### 2022-03-18 paper.tex
a paper i wrote a while ago

Having sat with this question for a bit longer (and reflected on the feedback you gave my
last essay), and considered the plays I did and didn't like this semester, I think:
theater because it can be specific where other art forms must be general. Specific in its
language, specific in its mode, specific in its content. Specific to its place, specific
to its time, specific to its audience. I'd like to talk about specificity and the
exploration it allows in four plays we experienced: \textit{Roma Armee}, \textit{The New
  Gospel}, \textit{Murmel Murmel} and \textit{Black Bismarck}.

My process in writing this essay was to first select the four plays from our class which
most affected me, the four about which I still had thoughts I needed to share, and then to
examine what it was about those plays that affected me. During that process, I settled on
the realization that the plays which moved me were the ones which spoke to my own lived
experience, and the plays which frustrated me were those that spoke to an experience I had
not lived. Having come to that realization, I think I could go back and talk about a
different set of four plays which might make a stronger argument; I might substitute
\textit{TANZ} in place of \textit{Murmel Murmel} or \textit{The New Gospel}. But I've
already written my four reflections, and I've come back to the top to further preface them
like this.

% I'd like to note, before we begin, that throughout the course of this semester I got my
% hands on Julia Jarcho's \textit{Writing and the Modern Stage}, and it's shaped the way I
% think about text, narrative and drama.

Much of my engagement with these plays is going to be in describing the conceptual spaces
they construct and inhabit. I contend that all art, and in fact all communication, is a
process of building a conceptual space and populating it with ideas, or from another
perspective, building a conceptual space which is capable of expressing the ideas you're
interested in. Math classes do this, computer programs do it, romance novels do it, and
plays do it. Each of these works builds upon an existing space specific to the person
consuming it, a space which is informed by expectations and past experiences such as those
with other works in the same medium or genre. For example, the latest Marvel movie can
drop quite easily into the space I bring with me because it was defined by the thirty
Marvel movies I've already seen, whereas a new HBO show must first inform me of the stakes
and the rules because I don't know whether to expect \textit{Curb Your Enthusiasm} or
\textit{Westworld}. The unique thing about theater is that a precise understanding of
one's audience allows a writer or performer to assume more of a shared foundation than is
possible for artists in other media.

\clearpage{}

\paragraph{\textit{Roma Armee}} is both the play that I enjoyed the most out of any we
watched, and the least interesting among those I'll discuss here. By this I mean, it is
the most readily recognizable as a play, albeit not a drama, and it is easy to recognize
both the play's goals and the methods it uses to achieve them. In terms of conceptual
spaces, \textit{Roma Armee} seemed to occupy the already-extant space of bourgeois
political theater, and used it to construct some ideas about how to perceive Romani
people.

Or at least, that was my experience. Based on our discussion in class, I gather that many
of my peers did not find \textit{Roma Armee} as straightforward as I do. On closer
reflection, my specific background prepared me to understand \textit{Roma Armee} more than
my classmates:

\begin{itemize}
\item As a practicing reform Jew with meaningful religious education, I have spent long
  enough studying the Shoah that my understanding goes beyond merely the unexplained
  deaths of Jews, but includes both knowledge of the propaganda which allowed the Nazis to
  kill all those people, and an understanding that Jews were not unique, but were the
  largest among several groups targeted by the Nazis, alongside the Roma. All this to say,
  I'd been taught, at least vaguely, who the Roma are, and what had been done to them.

\item Growing up, I was into superheroes, especially the X-Men. I was familiar with
  Magneto's history (and had spent some time critiquing the idea \enquote{what if the
    Nazi's victims actually were dangerous supernatural beings?}), and I remember being
  quite upset when the MCU incarnation of Wanda and Pietro Maximoff were generic eastern
  Europeans instead of Roma.

\item I spent enough time in Sweden (which is basically upstate Germany) to have a sense
  of how my cultural and historical awareness compared to the norm: especially politically
  active Europeans would have a deeper understanding of the history than me, as would
  directly affected minorities, but most had only a passing familiarity with the idea of
  the Roma.
\end{itemize}

I believe that \textit{Roma Armee} relied on all of this knowledge, and I also believe
that, pretty reliably, every person in the Gorki's audience for that performance would
have had it, although the background that would've given it to them might have been
different. The superhero thing was probably the least understood among the audience, and
it's also the thing most explicitly explained by the show.

\clearpage{}

\paragraph{\textit{The New Gospel}} doesn't fit my vision for theater. There seemed to be
some confusion about how to classify it, so let me say: this was a film. Reading Milo
Rau's manifesto, I can see how he ended up as a filmmaker. The last three of his points
are devoted to making theater transportable to places other than where it was conceived,
and to conceiving plays in places other than where they will be performed. The logical
result of these goals is film.

\begin{enumerate}[label={Command \arabic{*}}, start=8]
\item \label{raucommand:eight} requires that all plays be transportable in a car or
  van. Transporting a film requires only a projector and a screen, or even better, a
  simple digital transfer.

\item requires that plays be conceived in conflict or war zones (or performed there, but
  that is less interesting, and really just an extension of
  \ref{raucommand:eight}). \textit{The New Gospel} shows us that Rau's goal is to portray
  these conflicts to otherwise insulated audiences. Bringing stories from a conflict zone
  and staging them in Berlin does less to break that insulation than bringing people from
  the conflict zone and literally placing them onstage in Berlin, which in turn does less
  than recording images and sounds and recreating them wholesale.

\item extends \ref{raucommand:eight} by requiring that works actually be transported, not
  just that they be transportable. And it is here that Rau loses what I love about the
  theater: his works can no longer laser-target the expectations, experiences and
  preconceptions of an audience full of Berliners, because they also have to be applicable
  to the expectations, experiences and preconceptions of Parisians, Luxembourgers and
  residents of Istanbul (or wherever the Ghent decides to tour).
\end{enumerate}

With \textit{The New Gospel}, Rau set out to break the norms and expectations of theater
to align with his manifesto, and wound up creating a film. A very beautiful film,
certainly, but one that lacked the theater's ability to speak to its audience in their own
language, or to cooperate with its audience in constructing a shared language. We can see
this, in part, because it was the least referential of all the works I'll discuss: the
only cultural artifact referenced was the passions, which has got to be the most
universally known story there is, at least in Europe and the Americas.

\clearpage{}

\paragraph{\textit{Murmel Murmel}} makes this list because I can see that it was specific
to someone's background, but I did not have it. From our discussion, I can see that some
of my classmates did, though, and I am fascinated by the contrast between my experience
and theirs. This play seems to have depended on a framework of movement and physicality
which was present for the dancers and some of the actors in our class, and I suppose for
people with that framework, the things happening onstage communicated ideas in a way
analogous to the words and images in \textit{Roma Armee} or \textit{Black Bismarck} did to
me. For me, though, \textit{Murmel Murmel} was like reading a book in a foreign
language. Or maybe like being given the instruction manual for an IKEA shelf but having to
buy the parts from a Home Depot, and what you end up with just isn't right.

For the purposes of this assignment, I suppose I can let \textit{Murmel Murmel} serve as a
counterpoint, an exemplar of the downside of the hyperspecificity I'm describing. When a
play is for a specific group of people, and it depends on those peoples' foreknowledge,
any outsiders will find it more or less unintelligible. In addition to being better able
to predict its audience relative to film or other media, the theater is also, in my
opinion, more \enquote{allowed,} culturally, to be exclusionary in this way.  I think back
to some discussions I had about the film \textit{The Green Knight}. I loved the movie, but
I think I was only able to love it because I was used to post-dramatic theater, where plot
is not the guiding force. My friends and family who lacked that background did not like
\textit{The Green Knight}, and moreover, they seemed angry that there were movies in
theaters which weren't for them, which depended on a conceptual framework they didn't
have. For all I complain about \textit{Murmel Murmel}, I was certainly not surprised that
I didn't get it; that's just par for the course in the theater.

\clearpage{}

\paragraph{\textit{Black Bismarck},} more than anything else we've seen in this course, is
my vision for post-dramatic theater. I was not within its crosshairs, nor was I quite as
close as I was for \textit{Roma Armee}, but I was certainly much closer than I was for
\textit{Murmel Murmel} or \textit{TANZ}. My language requirement in German left me at
least aware of the linguistic and cultural phenomena the play was responding to.  I
imagine the actual audience, when shown the slideshow of all the monuments to Bismarck,
would have had a realization like, \enquote{huh, I walk past these all the time without
  ever looking too closely or questioning why they were there. I guess I never noticed
  just how prevalent Bismarck monuments are,} but since COVID spoiled my study abroad
plans, I never actually walked past any of the statues. I had seen \textit{Ghostbusters},
but I had to be reminded of the specific context of Stay-Puft's appearance; I had
forgotten about \enquote{clear your mind!}

But after investigating each of these things, I think I can imagine how \textit{Black
  Bismarck} affected its audience, and I am impressed. It seems likely to me that much of
the audience had traveled via that actual subway station, or at the very least one like
it, and had seen the statues that the white rabbit kicked over. A film could not have been
sure of that; filmmakers must contend with the possibility that the same version of
products will be shown in theaters across the country or the world, to people of wildly
different backgrounds, and even in informal settings like living rooms and
airplanes. (Ironically, that's exactly our experience with \textit{Black Bismarck}.)

But (when performed live), \textit{Black Bismarck} could assume those things, and it did,
to great effect. To an actual Berliner, who spoke German as a relatively uncritical
first-language speaker, who experienced the city and its monuments as the background of
everyday life, this play must have masterfully invited (or forced) them to step back and
to critically reexamine their world.

\clearpage{}

For my money, the advantage of theater's specificity is that plays have to do less work
establishing a shared context before they can get to something new and interesting. They
can be a scalpel instead of a chainsaw. If that's not enough for you, then in this space,
feel free to imagine a conclusion about how great theater is which uses the phrases
\enquote{refreshing contrast,} \enquote{intimate,} \enquote{mass-market entertainment,}
\enquote{globalization,} and \enquote{more important than ever.}

>>>>

What I propose is not, strictly, a unit of study, but rather, a focus that I would like to
see incorporated into the units already in your class. I find myself fascinated by the
economic realities of performance art, and the ways different systems of funding affect
the works produced. I was struck by this realization during a class from Brandon Woolf and
Andy Goldberg a few weeks ago, when we discussed Berlin's \textit{Freie Szene}. The norm
in Germany is for each theater to permanently employ an ensemble, who are paid a stable
salary by the state. The \textit{Freie Szene}, in contrast, consists of independent
troupes who are funded by government grants on a per-show basis. On the one hand,
\textit{freie} groups exist in a much more precarious space, where a series of rejected
proposals can leave them missing rent or collecting welfare. On the other hand, they enjoy
significant freedoms (hence the name) because of their disconnect from the institutional
hierarchies which employ more traditional ensembles. Notably, it is challenging for a
director working with a salaried ensemble to justify bringing in outside collaborators, so
temporarily hiring subject matter experts to collaborate on particular projects has become
a hallmark of the \textit{Freie Szene}. We got to see this play out onstage: \textit{Roma
  Armee}, a play interested in staging Romani performers, wound up with a cast that was
roughly half Romani, because the Gorki's ensemble only employed a handful of Romani
people. Rimini Protokoll's \textit{100\% Berlin}, on the other hand, invited and placed
onstage 100 Berlin residents. (In truth, I'm not sure if they were paid or volunteers, but
still.) I am interested in applying this same lens to the cultures of performance art we
discussed through the semester.

Throughout this essay, I will perform a relatively shallow investigation of various
economies of performance. I will compare these economies and their results to the
performance economy most familiar to me, Broadway (and American drama more broadly, as I
grew up in Minneapolis), but note that this is merely an artifact of convenience. My goal
is not to privilege Broadway as the optimal performance economy to which all history has
led; in fact, I think it's actually a pretty terrible version, and I find many of the
economies I'll discuss much more compelling.

Also note that most of what I write here is my own analysis based on my background
knowledge and what we discussed in class, along with a non-zero but frankly insufficient
amount of research. I caution you that what I write below will likely not reflect
scholarly consensus, because I'm not sure what the scholarly consensus is.\footnote{I
  blame this at least partially on the focus on novelty and reinterpretation in the
  humanities, which the computer scientist in me is unused to. In my discipline, it's not
  uncommon for the academic community to come to an agreement about the right way to do
  something, and then to stop worrying about it and move on to other things. This does not
  appear to be the case in theater studies or classics.}

\section*{5th century B.C.E. Athens}
% case study: greek drama

%% theater as a public service, performed by soldiers

Our class, like every theater history class, started with the Greeks, and so I suppose I
should, too. My understanding, based largely on a course I took with Peter Meineck to
satisfy my pre-1900s theater requirement, is that Athenian drama, was written by private
citizens for prestige, and performed by enlisted soldiers as part of their duties to the
city. I've also heard it proposed that wealthy Athenians would fund the playwrights, but
this was again with the aim of social prestige, not in exchange for artistic control or
seeking some return on investment. It's difficult to make broad statements about the
effects of this cultural organization on its products, as we have access to only a small
subset of the plays written by the three most successful playwrights, and none by any of
their myriad competitors. Still, we can trace some through-lines.

Athenian dramas tended to be quite specific to the year they were written and staged; I
think of \textit{Ajax} commenting on the rising tensions leading up to the Peloponnesian
war, and of \textit{Oedipus Tyrannus} responding to Pericles' death in the Plague of
Athens. Hyper-specific plays like this are discouraged in many other cultural economies,
including our own, by a desire that scripts be reusable and accessible. A Broadway show
which spoke deeply to its opening-night audience, but within a year started to alienate
all newcomers would be considered a failure, but the Athenian model allowed plays to do
just that.

Being performed as a public service, without a direct financial incentive like ticket
sales, seems to have encouraged Athenian playwrights to focus more on exposing their
audiences to new ideas or perspectives, and less on leaving them entertained or happy,
than the Broadway scene. I think once again of \textit{Ajax}, this time the ambiguous
light in which it portrays Ajax, one of Athens' founding heroes; Odysseus, arguably the
most beloved mythical hero; and Athena, the patron goddess of Athens. I'd love to tell you
what the Greek public thought of \textit{Ajax}, but all I can tell you is that there is no
record of its having won the City Dionysia. We do know that some very challenging and
provocative plays won first place, including Aeschylus' \textit{Persians}, which invited
sympathy for a hated enemy against whom Athens had quite recently fought and won a brutal
war. This suggests to me that the festival's convoluted judging scheme managed to select
for some definition of didacticism and artistic integrity, rather than pure entertainment
value.

\section*{Imperial Rome}

Roman theater, as I understand it, is best summarized by the phrase \enquote{bread and
  circuses.} Roman plays were funded either by the empire itself or by the patricians, and
served (with a range of conscious forethought) to placate and distract the plebeians and
to maintain the social hierarchy. At least as the post-Roman Christians tell it, this
meant the Romans' theater was gaudy, excessive and meaningless. Their main contribution to
drama as an art was in importing, preserving and re-performing the works of other
cultures; every surviving Greek play comes to us through the Romans. If I wanted to be
particularly political, I might set aside some time to draw parallels with modern America,
with the message that empire does not promote compelling art. That essay sounds like a
bummer, though, so instead I'll move on.

\section*{Medieval religious drama}

Frankly, I'm not that interested in this era, which is just as well because there's very
little to say about it. Medieval European history is characterized by the Church's
overwhelming influence over art, culture and day-to-day life. The Romans' excesses led the
Church to reject theater for the better part of a millennium, and when the art does
re-emerge, we see morality plays like \textit{Everyman}. It seems we actually know very
little about \textit{Everyman}; we don't know who wrote it, who first performed it, where
it was performed, or who paid for any of these things. I do feel safe to assert, though,
that the church was directly involved, either in a positive role by sponsoring the play's
creation or a negative one in restricting its content. Comparing this period with Broadway
theater is difficult, as our society is so secular compared to medieval Europe.

\section*{Shakespeare and the renaissance}

It's in the renaissance that we first see a theater economy similar to the one I live with
today, and by Shakespeare's time in the Elizabethan age, the parallels are
remarkable. Playwrights were commissioned by wealthy patrons, especially the monarchs, and
used that funding to support initial performances. They earned profit from ticket sales,
and used the prestige of popular shows to attract patrons for future works. The notable
difference compared to Broadway is that Shakespeare had no obligation to pay back his
patrons' gifts; they were not investors. But the model is fundamentally the same:
monarchs, aristocrats and other people of wealth used financial involvement to dictate the
content of new plays, and ticket sales provided an incentive to produce popular,
crowd-pleasing shows.

Patrons in this era used drama as a means of political conflict and propaganda. On the
world stage, this takes the form of the monarchs commissioning plays which glorify,
historicize and mythologize England's past. We can't actually say for certain which of
Shakespeare's plays were commissioned by Elizabeth or James, or at least, I can't find a
source which justifies the claim, but we do know that a disproportionate number of
Shakespeare's plays are devoted to England's frankly very boring history, and that the
monarchs were his primary patrons. More interesting, on a local level, we see at least one
instance of an internal dispute using a Shakespeare play to justify an attempted coup: the
Earl of Essex famously hired Shakespeare's company, the Chamberlain's Men, to perform
\textit{Richard II} in the hopes of swaying public opinion towards the would-be usurper
against Elizabeth. (It didn't work.)

The popular incentive of ticket sales mostly shows through in Shakespeare's dirty jokes
and reliance on convention. You've already read my take on the dirty jokes, though I'll
point out that they show up not only in \textit{Romeo and Juliet} but also in otherwise
serious tragedies; \textit{Theater Histories} cites the porter scene in
\textit{Macbeth}. In addition to saucy entertainment, audiences expected certain
conventions, especially in terms of plot and style. Comedies had to end with a marriage,
and tragedies with a death; \textit{Theater Histories} here points out that \textit{As You
  Like It} and \textit{Hamlet} had four each of the appropriate ending. Jeremy Lopez has a
whole book about the conventions expected by theater audiences at this time, and the way
playwrights (especially Shakespeare) sought to meet those expectations; he highlights puns
and wordplay, asides and soliloquy, and rapid tonal shifts to keep things exciting. Other
than the specific plot conventions, which have changed to accommodate a new zeitgeist, I
think this whole paragraph could apply to Broadway musicals (dramas tend to be more
experimental and less constrained); they're filled with crowd-pleasing plot moments like
the long-awaited kiss at the end, dirty jokes, songs which illustrate a character's
internal monologue, and pretty much every other device Lopez describes.
